Interactive Visual Analysis of Image-Centric Cohort Study Data

### 2014-06-17 draft.tex
Worked on the Application section

Morphometric features are derived to allow for statistical comparison of the tissue, which incorporates mostly positions, diameters, volumes and relative distances and alignment to other structures.
% 
% - Data Format\\
% - Missing Data\\
% - Registration of Image Data (VMV'12)\\
% - Extraction of Image Parameter
\subsection{Analysis Workflow}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/InteractionLoop}
 \caption{Detailed \emph{IVA Loop} as extension from Figure~\ref{fig:WorkflowComparison}.
 %
 Usually starting with a selection of a feature of interest (user-driven or via data mining techniques), the data are mapped using a visualization technique appropriate for the selected data types.
 %
 The data are visualized in the range and domain space, which can then be brushed, yielding new groups, to be investigated using further features.
 %
 Note that adjacent steps are directly connected via feedback loops allowing for an iterative refinement and giving as much freedom to the user as possible.}
 \label{fig:InteractionLoop}
\end{figure}
%-------------------------------- Figure --------------------------------
\begin{figure*}[htb]
 \centering
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 % \includegraphics[width=3.5in]{figures/visualization}
 %\includegraphics[width=1\textwidth]{figures/visualization}
 \includegraphics[width=1\textwidth, resolution=300]{figures/visualization}
 \caption{(Left) Screenshot from the front-end, which is divided as follows: (a) The sidebar containing all features as well as the groups defined in the analysis process; (b) the canvas area where features can be added via drag and drop and the visualization can be chosen automatically according to the data type; (c) the interactive pivot table showing the exact numbers for each displayed feature combination. The data displayed is used to analyze the lumbar spine. Features can be added freely on the canvas via drag and drop. Dropping the \emph{gender} parameter on the already plotted \emph{body size} container creates a mosaic plot combining both features (right).
 }
 \label{fig:visualization}
\end{figure*}
%-------------------------------- /Figure -------------------------------
Our proposed \emph{IVA} workflow knows three major steps as illustrated in Figure~\ref{fig:InteractionLoop}: Feature selection, visualization and brushing.
%
A hypothesis-driven analysis usually starts with the selection of features.
%
A feature selection from a shape-based clustering, which creates shape groups.
%
\emph{Hypothesis generation} with focus on image data starts with a shape-based clustering.
%
\com{Multivariate analysis using the adjacency matrix may have also be explained here, depending on the Application chapter.}
%
The feature is mapped using an automatically chosen visualization appropriate for its data type (described in detail in the following section).
%
The visualization techniques have to combine both image-and non-image data in order to set domain and range data in relation to each other.
%
In our system, the visualization can either be brushed or new features can be added to the analysis. 
%
Brushing methods are subdivided using the previously described \emph{IVA} patterns.
%
Brushed regions are treated like features as they divide the subject space just like categorical features.
%
Selecting features also triggers a \emph{multivariate analysis} using contingency values (described in the following section) to highlight features associated with selected features.
%
A sample workflow using interaction and visualization techniques described in the next section can be seen in Figure~\ref{fig:visualization}.

@@

\subsection{Implementation}
\begin{figure}[htb]
 \centering
 \label{fig:technologies}
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 \includegraphics[width=3.5in]{figures/technologies}
 \caption{The front-end solution (left) uses state-of-the-art web technologies such as \emph{HTML5}/\emph{CSS3}, \emph{WebGL} and \emph{SVG} to display the data.
 % 
 The \emph{NodeJS} based back-end (right) stores all image and non-image data and transfers it to connected clients.
 %
 All computation-heavy operations, such as calculation of mean shapes or distances, as well as statistical processing are done by the server to keep hardware requirements of client systems low. 
 %
 Client-server communication is accomplished via the \emph{Websocket} protocol.
 }
\end{figure}
In this section we discuss how we implemented the presented methods using free, open web standards.
%
To provide a fast communication loop between method development and expert input, we decided to rely on modern web technologies.
%
In addition to the obvious advantages of web technologies, the following aspects are crucial for our work:
\begin{itemize}
%	\item No additional software needs to be installed, most people use a decent state-of-the-art web browser, even on mobile devices.
	\item The client-server structure allows for employing heavy computation on a capable server machine and transfering results to the client.
	\item Since image data for several thousand subjects requires hundreds of gigabytes disk space, it can remain safely on the server and elements can be transferred on demand.
	%
	High confidentiality standards of the data can be met by restricting access via an account system.
	\item Recent developments in \emph{WebGL} applications running in browsers with near-native performance push the development further into to the web resulting in many open source libraries, which are well documented, rich in examples and driven by active communities. We use \emph{WebGL} for rendering shape information.
%	\item Web technologies are free.
\end{itemize}
These advantages do not come without drawbacks.
%
Due to missing ports of sophisticated and specialized libraries/languages, such as the \texttt{Visualization Toolkit} (\texttt{VTK})\footnote{Developed by Kitware Inc; \url{http://vtk.org}} or \texttt{R}\footnote{Open Source; \url{http://r-project.org}} for statistics, many standard methods need to be written from scratch for the web.
%

@@

Shape distance is mapped onto color.
%
For dichotomous variables, the color codes distances between mean shapes of the two groups, for variables with more than two manifestations it encodes the distance to the global mean shape of all subjects (see Figure~\ref{fig:hypothesisfree}).
\\\\
Shape-based clustering is carried out via agglomerative hierarchical clustering of the spine canal centerlines (recall Section~\ref{application:Data Preprocessing} and \cite{Klemm2013VMV}).
%
Since the ``correct'' number of clusters in a given group is unknown, an estimate is computed by means of the knee/elbow method \cite{Salvador2004}.
%
The method has proven to produce sound results on a preliminary data set and was able to reproduce textbook knowledge \cite{Klemm2013VMV}.
\subsection{Domain Experts, Setup and Procedure}
%
%Lam et al. divide evaluations of Visual Analysis systems in two categories, 
%
\com{Anmerkung von Steffen einbauen: Evaluierung geht um fachliches Feedback - Wissen über epidemiologie UND rücken und wissen um die Daten - deswegen schwer ganz viele Experten ranzuziehen}
Guided by the work of Lam et al \cite{Lam2012}, we conducted an investigation of \emph{Visual Data Analysis and Reasoning} (VDAR).
%
This approach aims to characterizes the ability of a system to explore data, discover knowledge, generate hypotheses and help formulating decisions.
%
Since it is hard to quantify when these goals are achieved, Lam et al. suggest case studies for the VDAR analysis.

We conducted a qualitative case study with two epidemiological domain experts.
%
Both of them are practicing clinicians with focus on epidemiological research.
%
The first clinician is an specialist in internal medicine and also the leader of the SHIP.
%
One clinician is an radiologist and responsible for the image acquisition of the SHIP and has a deep understanding on the available image data.
%
\paragraph{Setup.} To bypass the large geographical distance between us and the domain experts and to take their busy schedule into account, the case study was done completely web based.
%
Since one major reason for the focus on web technologies, no software needed to be exchanged.
%
A website link was send to the users, which they could open with their browser of choice and start using the tool.
%
To make the user input comprehensible, we observed their monitor via screen sharing.
%
Communication was enabled via webcam supported voice over ip.
%
The total setup time including installing the screen sharing application was about five minutes.
%
Video-recordings of the sessions made it possible to evaluate the sessions later on.
\com{TODO: Name der Screensharing Software mit angeben (Teamviewer)?}
\paragraph{Procedure.}
At first, controlled mouse and keyboard of the clinicians pc and demonstrated the basic functionalities of the system.
%
This included the Heatmap, the correlation view, how to introduce parameter visualization, how the color coding of differences works and the pivot tables.
%
As they understood the concepts, we handed over the mouse and keyboard control and only observed from this point on.
%
The epidemiologists were given two tasks: one hypothesis-free analysis of the data and one starting with an assumption.

@@

\paragraph{Analysis 1.}
%
The latter was chosen first by both domain experts.
%
%They pointed out the high potential they see in this visualization, because it allows to look at the data in an unbiased way.
They pointed out the benefits of the possibility to look at all variables in the context of each other.
%
To cite the radiologist, it "illumunates the black box of the data", making it possible to look at the data in an unbiased way.
%
Due to the dense mapping of information in the contingency matrix it was suggested to make this visualization full screen.
%
The radiologist was looking for correlations with shape related features in the data, finding that they correlate strong with each other as well as with leg pain, age, body height and hormone replacement therapy.
%
The internist inspected features which are not derived from the images, such as parameters strongly related with age, such as income, blood fat values or number of born children.
%
This also shows, that even though the visualization may treat every feature equally, the domain experts is more likely interested in features they are familiar with and have personal clinical experience with.
%
%Many correlations observed in this view

After this initial overview, the expert performed a \emph{multivariate analysis} by introducing features, such as age, waist circumference, weight or lumbar spine centerline curvature as bar charts views into the canvas area and selected subgroups to see how they are distributed and if they could observe unusual behavior in the mean-shapes.
%
+ Cramer test 
\com{Allgemein: Workflow noch einmal grob beschreiben - Overview first, details on demand}
%
\paragraph{Analysis 2.}
\begin{figure}[htb]
 \centering
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 \includegraphics[width=3.5in]{figures/hypothesisfree}
 \caption{(a) Clustering result of all subjects. The bar chart height indicates the number of subjects in the cluster. The difference to the mean shape is color-coded, whereas blue represents no difference and red a large difference. (b) A pivot table relates cluster results to gender and employment status. Cluster 8 is striking because of the high proportion of women, while cluster 2, 3, 4 contain almost only men.}
 \label{fig:hypothesisfree}
\end{figure}
\com{ToDo: Rewrite}
%
The internist started also with a \emph{multivariate analysis} using the contingency table, but found no associations he was interested in.
%
Therefore, he applied the \emph{local investigation} pattern by a shape grouping step using shape-based clustering via dragging the "All subjects" feature into the canvas area, triggering the shape clustering.
%Shape-based hypothesis-free exploration starts with the \emph{local investigation} pattern by a shape grouping step using shape-based clustering.
%
The results for this step can be seen in Figure~\ref{fig:hypothesisfree} (a).

Cluster 9 represents subjects with average shape.
%
Other shapes differ with respect to size, such as cluster 2, 3, 8, 10 where the last one also represents a more straight spine, which is usual for subjects with larger body size.
%
Cluster 4, 5, 7 contain outliers, characterized by their unusual shape and small number.
%
Cluster 8 has the second largest number of elements and was therefore of special interest.
%
The user opened the contingency pane to performing a \emph{multivariate analysis} by looking at \emph{Cram\'{e}r's V} contingency values of cluster~8.
%
This revealed associations of this group with employment status, gender, body size, age, thyroid nodules and blood fat value.
%
Another \emph{multivariate analysis} using a pivot table set gender and employment status in relation to clustering affiliation\add{.} 
%
\add{It s}hows that cluster~8 contains mostly women and also has a larger unemployment rate (see Figure~\ref{fig:hypothesisfree}), while the overall employment rate of women and men in the data set is almost exactly 50\%.
%
While all observed features seem to be plausible associations related to back pain, the values indicate that cluster~8 contains subjects with chronic back pain radiating to the legs.
%
Metabolic parameters, such as blood fat and blood sugar, are also possibly associated features.
%
The employment status is a feature relating to many different lifestyle factors such as income or nutrition as well as age and might act as confounder.

% TODO: Contingency Matrix or Adjecency Matrix
\com{00.44 Clustering-Ergebnis Weiblich und Arbeitspensum Alter}

\subsection{Case 2: Hypothesis-based analysis}
If the user proposes a hypothesis about a relation between a non-image feature regarding shape, the workflow slightly differs from the hypothesis-free analysis.
%
The starting point follows the \emph{feature investigation} pattern, where a feature of interest is selected by dragging it into the canvas area and viewing the subject's distribution as well as their shape differences.
\paragraph{Analysis 1}
The radiologist performed a \emph{feature investigation} by dragging the age variable into the canvas area, observing that age-bins 36-52 years and 52-68 years are nearly identical to the global mean, while 20-36 years and 68-84 years differ in size.
%
Performing a \emph{multivariate analysis} by selecting only female subjects via the additionally added gender feature it was observed, that distance was closest to subjects of 68-84 years old and largest to 20-36 years old.
%
The expert argued that this is most likely due to the shrinking of the spine with increasing age, making it more similar to the female average spine which is smaller than the global average \com{(00.29)}.
%

\paragraph{Analysis 2.}
Hypothesis: "Back pain is associated with age and therefore there differences in the lumbar spine shape are observable." \com{TODO Image Mosaic Plot Age, Backpain}.
To validate this hypothesis, a \emph{multivariate analysis} was performed by introducing the dichotomous feature "\emph{Did you experience backpain in the last three moth?}" together with the age as mosaic plot by dropping both features on the canvas area.
%
The experts were not able to observe the expected effect in the visualization.
%
Reasons for this are twofold.
%
For one, age influences the lumbar spine shape very heavily, as seen in Fig~\ref{?}, while the differences between subjects with and without back pain are small.
%
The major difference seen in the visualization are therefore presented by the age feature, masking differences via the back pain parameter.

The second explanation is the commonality of back pain in our society.
%
As seen in Fig~\ref{?}, subjects reporting back pain are even the majority, which makes it difficult to extract parameters which reliably describe back pain.
%
A \emph{local investigation} using the contingency table showed a strong association between back pain with gender and body height.
%
Body height was explained to be a confounder for gender, since female subjects smaller on average than male subjects.
%
The analysis solely based on shape only accentuated gender differences, not in back pain.
%This makes it hard to analyze shape related features for this disease
%
A next step would be to use a variable describing the strength of back pain and only analyzing strong back pain!
%

\com{ToDo: Ability to better create sub groups would be useful as well as better segmentation}
%gender, use of pain medications. this would be good for future work, providing a better way to subdivide the data set and construct groups. The presented method is very good to gain an overview

@@

\paragraph{Analysis 3}
\com{ToDo: Needs to be rewritten}
\begin{figure}[htb]
 \centering
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 \includegraphics[width=3.5in]{figures/hypothesisbased}
 \caption{(a) Dichotomous questionnaire answer to ``\emph{Did you experience back pain in the past three months?}''.
 %
 Mean shapes between the groups show no difference.
 %
 (b) Shape-based clustering for all subjects who suffered from back pain yields 6 groups. 
 %
 Note that the difference in subject count is due to the missing shape information for some subjects.
 %
% The total count of subjects in (b) is smaller 
 }
 \label{fig:hypopthesisbased}
\end{figure}
%
In our use case, epidemiologists were interested in the dichotomous questionnaire answer to ``\emph{Did you experience back pain in the past three months?}''.
%
The mean shapes of the resulting visualization show no difference between the two groups (see Figure~\ref{fig:hypopthesisbased} (a)).
%
Either there are no differences or the variance information was lost in the mean shape calculation.
%
Since the focus is on subjects suffering from back pain, the clustering results of these subjects are then drawn into the canvas area, yielding six clusters as seen in Figure~\ref{fig:hypopthesisbased} (b).
%
Cluster 5 stood out for having a so-called \emph{hyperlordosis}, a strong curvature of the lumbar spine associated with back pain.

\emph{Multivariate analysis} using \emph{Cram\'{e}r's V} contingency values highlighted relationships of this cluster with joint degeneration, meat eating habits, preoccupation, back pain, neck or shoulder pain and waist circumference.
%
Since the prior selection only yields subjects that report back pain, the pain indicators specify the pain localization for the subjects.

It is well known that overweight is a risk factor for back pain.
%
While the BMI is a key figure for assessing height and weight of a subject, it does not tell anything about how the weight is distributed in the body.
%
Our epidemiological collaborators were interested in the correlation with waist circumference presented by this group.
%
Our finding follows the recent trends indicating that BMI is not a good measure for assessing the body shape, since healthy weight is dependent on many other measures \cite{Ahima2013}.
%
It indicates that the waist circumference rather than the BMI interacts with unusually shaped spines for subjects with lumbar back pain.
%
The influence of the parameter is now in the focus of further analysis.

\com{00.46 Second: Analysis of Female subjects via clustering, selction of cluster 2, analysis using Body weight and Age. Clustering Result if 1 gives large distance between other subjects. 00.49 How are can the differences be explained?}
\com{00.46 Feedback Body-Size/Body weight. Some features are sparse, suggestion was made to use quantile/quintile, resulting in equally occupied results. This also depends on what the user is intersted in - for clinical use the outliers may be interesting (57.30). (58.30) From perspective of a radiologists, quartile and quintiles make sense when establishing reference values for diseases; clinical relevant parameter, einflussfaktoren oder risikofaktoren zu identifizieren, je nach dem welche hypothese man verwendet.}

>>>>

Morphometric features are derived to allow for statistical comparison of the tissue, which incorporates mostly positions, diameters, volumes and relative distances and alignment to other structures.
% 
% - Data Format\\
% - Missing Data\\
% - Registration of Image Data (VMV'12)\\
% - Extraction of Image Parameter
\subsection{Analysis Workflow}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/InteractionLoop}
 \caption{Detailed \emph{IVA Loop} as extension from Figure~\ref{fig:WorkflowComparison}.
 %
 Usually starting with a selection of a feature of interest (user-driven or via data mining techniques), the data are mapped using a visualization technique appropriate for the selected data types.
 %
 The data are visualized in the range and domain space, which can then be brushed, yielding new groups, to be investigated using further features.
 %
 Note that adjacent steps are directly connected via feedback loops allowing for an iterative refinement and giving as much freedom to the user as possible.}
 \label{fig:InteractionLoop}
\end{figure}
%-------------------------------- Figure --------------------------------
\begin{figure*}[htb]
 \centering
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 % \includegraphics[width=3.5in]{figures/visualization}
 %\includegraphics[width=1\textwidth]{figures/visualization}
 \includegraphics[width=1\textwidth, resolution=300]{figures/visualization}
 \caption{(Left) Screenshot from the front-end, which is divided as follows: (a) The sidebar containing all features as well as the groups defined in the analysis process; (b) the canvas area where features can be added via drag and drop and the visualization can be chosen automatically according to the data type; (c) the interactive pivot table showing the exact numbers for each displayed feature combination. The data displayed is used to analyze the lumbar spine. Features can be added freely on the canvas via drag and drop. Dropping the \emph{gender} parameter on the already plotted \emph{body size} container creates a mosaic plot combining both features (right).
 }
 \label{fig:visualization}
\end{figure*}
%-------------------------------- /Figure -------------------------------
Our proposed \emph{IVA} workflow knows three major steps as illustrated in Figure~\ref{fig:InteractionLoop}: Feature selection, visualization and brushing.
%
A hypothesis-driven analysis usually starts with the selection of features.
%
A feature selection from a shape-based clustering, which creates shape groups.
%
\emph{Hypothesis generation} with focus on image data starts with a shape-based clustering.
%
\com{ToDo: Multivariate analysis using the adjacency matrix may have also be explained here, depending on the Application chapter.}
%
The feature is mapped using an automatically chosen visualization appropriate for its data type (described in detail in the following section).
%
The visualization techniques have to combine both image-and non-image data in order to set domain and range data in relation to each other.
%
In our system, the visualization can either be brushed or new features can be added to the analysis. 
%
Brushing methods are subdivided using the previously described \emph{IVA} patterns.
%
Brushed regions are treated like features as they divide the subject space just like categorical features.
%
Selecting features also triggers a \emph{multivariate analysis} using contingency values (described in the following section) to highlight features associated with selected features.
%
A sample workflow using interaction and visualization techniques described in the next section can be seen in Figure~\ref{fig:visualization}.

@@

% TODO: This can be an Section of it's own? Or part of the design?
\subsection{Implementation}
\begin{figure}[htb]
 \centering
 \label{fig:technologies}
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 \includegraphics[width=3.5in]{figures/technologies}
 \caption{The front-end solution (left) uses state-of-the-art web technologies such as \emph{HTML5}/\emph{CSS3}, \emph{WebGL} and \emph{SVG} to display the data.
 % 
 The \emph{NodeJS} based back-end (right) stores all image and non-image data and transfers it to connected clients.
 %
 All computation-heavy operations, such as calculation of mean shapes or distances, as well as statistical processing are done by the server to keep hardware requirements of client systems low. 
 %
 Client-server communication is accomplished via the \emph{Websocket} protocol.
 }
\end{figure}
In this section we discuss how we implemented the presented methods using free, open web standards.
%
To provide a fast communication loop between method development and expert input, we decided to rely on modern web technologies.
%
In addition to the obvious advantages of web technologies, the following aspects are crucial for our work:
\begin{itemize}
%	\item No additional software needs to be installed, most people use a decent state-of-the-art web browser, even on mobile devices.
	\item The client-server structure allows for employing heavy computation on a capable server machine and transfering results to the client.
	\item Since image data for several thousand subjects requires hundreds of gigabytes disk space, it can remain safely on the server and elements can be transferred on demand.
	%
	High confidentiality standards of the data can be met by restricting access via an account system.
	\item Recent developments in \emph{WebGL} applications running in browsers with near-native performance push the development further into to the web resulting in many open source libraries, which are well documented, rich in examples and driven by active communities. We use \emph{WebGL} for rendering shape information.
%	\item Web technologies are free.
\end{itemize}
These advantages do not come without drawbacks.
%
Due to missing ports of sophisticated and specialized libraries/languages, such as the \texttt{Visualization Toolkit} (\texttt{VTK})\footnote{Developed by Kitware Inc; \url{http://vtk.org}} or \texttt{R}\footnote{Open Source; \url{http://r-project.org}} for statistics, many standard methods need to be written from scratch for the web.
%

@@

Shape distance is mapped onto color.
%
For dichotomous variables, the color codes distances between mean shapes of the two groups, for variables with more than two manifestations it encodes the distance to the global mean shape of all subjects (see Figure~\ref{fig:hypothesisfree}).
\\\\
Shape-based clustering is carried out via agglomerative hierarchical clustering of the spine canal centerlines (recall Section~\ref{application:Data Preprocessing} and \cite{Klemm2013VMV}).
%
Since the ``correct'' number of clusters in a given group is unknown, an estimate is computed by means of the knee/elbow method \cite{Salvador2004}.
%
The method has proven to produce sound results on a preliminary data set and was able to reproduce textbook knowledge \cite{Klemm2013VMV}.
\subsection{Domain Experts, Setup and Procedure}
%
%Lam et al. divide evaluations of Visual Analysis systems in two categories, 
%
%\com{Anmerkung von Steffen einbauen: Evaluierung geht um fachliches Feedback - Wissen über epidemiologie UND rücken und wissen um die Daten - deswegen schwer ganz viele Experten ranzuziehen}
Guided by the work of Lam et al \cite{Lam2012}, we conducted an investigation of \emph{Visual Data Analysis and Reasoning} (\emph{VDAR}).
%
This approach aims to characterizes the ability of a system to explore data, discover knowledge, generate hypotheses and help formulating decisions.
%
Since it is hard to quantify when these goals are achieved, Lam et al. suggest case studies for the \emph{VDAR} analysis by applying the think aloud protocol to understand the domain experts observations, inferences and conclusions when using the system.

We conducted a qualitative case study with two epidemiological domain experts who also co-authored this publication.
%
Both of them are practicing physicians with focus on epidemiological research.
%
One is an specialist in internal medicine and also the head of the SHIP, the other an radiologist and responsible for the MRI-data acquisition of the entire SHIP. %and has a deep understanding on the available image data.
%
(Their expertise and experience with the data in their day to day research make them aware of the problems associated with the analysis of medical cohort study data.)
%
\paragraph{Setup.} To bypass the large geographical distance between us and the domain experts and to take their busy schedule into account, the case study was done completely web based.
%
Since one major reason for the focus on web technologies, no software needed to be exchanged.
%
A website link was send to the users, which they could open with their browser of choice and start using the tool.
%
To make the user input comprehensible, we observed their monitor via screen sharing.
%
Communication was enabled via webcam supported voice over ip.
%
The total setup time including installing the screen sharing application was about five minutes.
%
Video-recordings of the sessions made it possible to evaluate the sessions later on.
%
\paragraph{Procedure.}
At first, we controlled mouse and keyboard of the participants pc and demonstrated the basic functionalities of the system.
%
This included the Heatmap, the correlation view, how to introduce parameter visualization, how the color coding of differences works and the pivot tables.
%
As they understood the concepts, we handed over the mouse and keyboard control and only observed from this point on.
%
The epidemiologists were given two tasks: one hypothesis-free analysis of the data and one starting with an assumption.

@@

\paragraph{Analysis 1.}
%
The latter was chosen first by both domain experts.
%
%They pointed out the high potential they see in this visualization, because it allows to look at the data in an unbiased way.
%They pointed out the benefits of the possibility to look at all variables in the context of each other.
Before, they were not able to to look at all variables in the context of each other.
%
%
To cite one expert, it "illumunates the data black box", making it possible to look at the data unbiased from assumptions.
%
The radiologist was looking for correlations with shape related features in the data, finding that they correlate with \emph{leg pain}, \emph{age}, \emph{body height} and \emph{hormone replacement therapy status}.
%
The internist inspected non-image-derived features, such as age-associated parameters like \emph{income}, \emph{blood fat values} or \emph{number of born children}.
%
This also shows, that even though the visualization may treat every feature equally, the domain experts is more likely interested in features they are familiar with and have personal clinical experience with.
%
Due to the dense mapping of information in the contingency matrix it was suggested to make this visualization full screen.
%Many correlations observed in this view

After this initial overview, the expert performed a \emph{multivariate analysis} by introducing features, such as \emph{age}, \emph{waist circumference}, \emph{weight} or \emph{lumbar spine canal curvature} as bar charts views into the canvas area and selected subgroups to see how they are distributed and if they could observe unusual behavior in the mean-shapes.
%
This pointed out problems with the used categorization method splitting numerical variables into equally sized bins.
%
If an feature contains outliers, such as waist circumference (e.g. subjects with morbid obesity) this approach leads to sparse categories, making it hard to calculate associations.
%
Solutions for this were discussed with the domain experts and are described in more detail in Section~\ref{Lessons Learned}.
%
%\com{Allgemein: Workflow noch einmal grob beschreiben - Overview first, details on demand}
%
\paragraph{Analysis 2.}
\begin{figure}[htb]
 \centering
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 \includegraphics[width=3.5in]{figures/hypothesisfree}
 \caption{(a) Clustering result of all subjects. The bar chart height indicates the number of subjects in the cluster. The difference to the mean shape is color-coded, whereas blue represents no difference and red a large difference. (b) A pivot table relates cluster results to gender and employment status. Cluster 8 is striking because of the high proportion of women, while cluster 2, 3, 4 contain almost only men.}
 \label{fig:hypothesisfree}
\end{figure}
\com{ToDo: Rewrite}
%
The internist started also with a \emph{multivariate analysis} using the contingency table, but found no associations he was interested in.
%
Therefore, he applied the \emph{local investigation} pattern by a shape grouping step using shape-based clustering via dragging the "All subjects" feature into the canvas area, triggering the shape clustering.
%Shape-based hypothesis-free exploration starts with the \emph{local investigation} pattern by a shape grouping step using shape-based clustering.
%
The results for this step can be seen in Figure~\ref{fig:hypothesisfree} (a).

Cluster 4 represents subjects with average shape.
%
Other shapes differ with respect to size, such as cluster 2, 3, 7, where the last one and cluster 5 also represent a more straight spine, which is usual for subjects with larger body size.
%
Cluster 1 and 6 contain outliers, characterized by their unusual shape and small number.
%
%Cluster 8 has the second largest number of elements and was therefore of special interest.
%
%To perform an automatic feature suggestion using a \emph{multivariate analysis} by looking at \emph{Cram\'{e}r's V} contingency values of in the contingency pane.
To get an overview over the suggested features, the user opened the contingency pane to perform a \emph{multivariate analysis} by looking at \emph{Cram\'{e}r's V} contingency values of all clusters, revealing a strong correlation with \emph{gender} and \emph{body size}.
%
Therefore another \emph{multivariate analysis} was carried out by the feature \emph{gender} to the canvas from the contingency pane and selecting all female subjects (Fig.~\ref{ToDo}).
%
Cluster 1 contained almost only female subjects.
%
Contingency values for this cluster revealed correlations with \emph{leg fatigue}, \emph{physical heavy work}, \emph{body weight}, \emph{dyspnoea} and \emph{headache intensity}.
%
%This revealed associations of this group with employment status, gender, body size, age, thyroid nodules and blood fat value.
%
Since it is an pain indicator, headache was of special interest and was further investigated by incorporating a pivot table setting \emph{headache intensity} in relation to cluster affiliation.
%
Cluster 1 subjects report a much higher chance for heavy headaches than other subjects.
%Another \emph{multivariate analysis} using a pivot table set gender and employment status in relation to clustering affiliation\add{.} 
%
%\add{It s}hows that cluster~8 contains mostly women and also has a larger unemployment rate (see Figure~\ref{fig:hypothesisfree}), while the overall employment rate of women and men in the data set is almost exactly 50\%.
%
%While all observed features seem to be plausible associations related to back pain, the values indicate that cluster~8 contains subjects with chronic back pain radiating to the legs.
%
%Metabolic parameters, such as blood fat and blood sugar, are also possibly associated features.
%
%The employment status is a feature relating to many different lifestyle factors such as income or nutrition as well as age and might act as confounder.
\\\\
The experts emphasized the importance of methods providing overview over the data for hypothesis generation.
%
With the presented \emph{IVA} approaches they were quickly able to confirm medical knowledge as well as elaborate new hypothesis.
% TODO: Contingency Matrix or Adjecency Matrix
%\com{00.44 Clustering-Ergebnis Weiblich und Arbeitspensum Alter}

\subsection{Case 2: Hypothesis-based analysis}
If the user proposes a hypothesis about a relation between a non-image feature regarding shape, the workflow slightly differs from the hypothesis-free analysis.
%
The starting point follows the \emph{feature investigation} pattern, where a feature of interest is selected by dragging it into the canvas area and viewing the subject's distribution as well as their shape differences.
% \paragraph{Analysis 1.}
% The radiologist performed a \emph{feature investigation} by dragging the age variable into the canvas area, observing that age-bins 36-52 years and 52-68 years are nearly identical to the global mean, while 20-36 years and 68-84 years differ in size.
% %
% Performing a \emph{multivariate analysis} by selecting only female subjects via the additionally added gender feature it was observed, that distance was closest to subjects of 68-84 years old and largest to 20-36 years old.
% %
% The expert argued that this is most likely due to the shrinking of the spine with increasing age, making it more similar to the female average spine which is smaller than the global average \com{(00.29)}.
% %
% 
\paragraph{Analysis 1.}
Hypothesis: ``\emph{Back pain is associated with age and lumbar spine shape}''.
%Hypothesis: "Back pain is associated with age and therefore there differences in the lumbar spine shape are observable." 
\com{TODO Image Mosaic Plot Age, Backpain}.
%
To validate this hypothesis, a \emph{multivariate analysis} was performed by introducing the dichotomous feature ``\emph{Did you experience back pain in the last three month?}'' together with the age as mosaic plot by dropping both features on the canvas area.
%
The experts were not able to observe the expected effect in the visualization.
%
Reasons for this are twofold.
%
For one, age influences the lumbar spine shape very heavily, as seen in Fig~\ref{ToDoA}, while the differences between subjects with and without back pain are small.
%
The major difference seen in the visualization are therefore presented by the age feature, masking differences via the back pain parameter.

The second explanation is the commonality of back pain in our society.
%
As seen in Fig~\ref{ToDoA}, subjects reporting back pain are even the majority, which makes it difficult to extract parameters which reliably describe back pain.
%
A \emph{local investigation} using the contingency table showed a strong association between back pain with gender and body height.
%
Body height was explained to be a confounder for gender, since female subjects smaller on average than male subjects.
%
The analysis solely based on shape only accentuated gender differences, not in back pain.
%This makes it hard to analyze shape related features for this disease
%
The next step is to use a variable describing the strength of back pain and only analyzing strong back pain!

The epidemiologists pointed out that they would like to see a more intuitive and fast way to select subgroups based on different features to make full use of the analysis capabilities.
%
We discuss this in detail in Section~\ref{Lessons Learned}.
%gender, use of pain medications. this would be good for future work, providing a better way to subdivide the data set and construct groups. The presented method is very good to gain an overview

@@

\paragraph{Analysis 2.}
Hypothesis: ``\emph{Back pain is related to lumbar spine deformation.}''.
%
The previously discussed analysis puts the suitability of the lumbar spine segmentation for analyzing back pain into question.
%
The results of the analysis lead to this analysis.
%
This hypothesis is analyzed by dropping the feature ``\emph{Did you experience back pain in the past three months?}'' into the canvas are, as seen in previous Fig.~\ref{ToDoA}.
%
Then, the features derived from the clustering results for both subject groups are dropped into the canvas area as well (Fig.~\ref{ToDoB} (a)).
%
While the clustering algorithm finds only three homogenous cluster close to the mean shape for subjects reporting no back pain.
%
The cluster analysis for back pain yields diverse clusters with various pathological shape classes.
%
Cluster~5 represents most of the subjects and is very similar to the global mean shape.
%
Cluster~1 and 2 present a \emph{hyperlordosis}, a strong curvature of the lumbar spine, while Cluster 3 and 4 present a more straight shape.
%
A \emph{multivariate analysis} using the pivot tables put gender and strong back pain in context to cluster affiliation (Fig.~\ref{ToDoB} (b)). 
%
It shows, that subjects in Cluster~1, 2 and 6 reporting strong back pain while at the same time they also have a considerably higher share of females.
%
To check for unusual correlations, the expert opened the \emph{Cram\'{e}r's V} contingency table.
%
It depicted strong association with \emph{body fat}, \emph{body weight} and \emph{blood pressure} (Cluster~1) \emph{alcohol consumption} and \emph{attentiveness disorder} (Cluster~2), and \emph{amount of sleep} (Cluster~6).
%
For the experts, these observations are a starting point for a number of new hypothesis about possible relationships, for example association between overweight and Cluster~1.
%
\\\\
In summary it can be stated that hypothesis-driven analysis leads to hypothesis generation by design of the tool.
%
It is not designed to statistically validate hypothesis like statistical processors.
%
It rather triggers the analysis of potentially associated features with a pathology of interest.
%Zusammenfassend kann gesagt werden, dass hypothesenanalyse und hypothesengenerierung durch das design des Tools hand in hand gehen, es istn icht dafür geeignet statistisch bis ins detail hypothesen zu validieren, sondern triggert eher die analyse potentiell assoziierter variablen, die später verfeinert werden.
%
%By introducing the \emph{gender} feature and selecting all male subjects it was shown, that Cluster 1, 2 and 6 pr
%\emph{Multivariate analysis} using \emph{Cram\'{e}r's V} contingency values highlighted relationships of this cluster with joint degeneration, meat eating habits, preoccupation, back pain, neck or shoulder pain and waist circumference.
